draft

### 2025-05-20 draft.tex
docs: Update draft text, convert to Oxford style

% Created 2025-05-19 Mon 18:10
% Intended LaTeX compiler: pdflatex
\documentclass{article}

@@

\begin{abstract}
Quantifying the contents of objects in images is a common challenge in biological imaging. The most widely used software to do so require significant manual intervention. Here we introduce our library cp\_measure, which provides programmatic access to the most widespread metrics to convert images and objects into features. We then demonstrate that the features are consistent to the standard ones and showcase tasks for which our tool is more suitable than the alternatives. Our tool opens the door to community-driven  development and expansion of bioimage analysis metrics and pipelines, increasing developer accessibility and reproducibility of the pipelines.
\end{abstract}
\section{Introduction}
\label{sec:org4c9ba67}
The field of biological imaging leverages data from cell microscopy images to obtain biological insights. Modern microscopes can capture a wisde range of information, such as illumination and fluorescence. When coupled with fluorescence dyes or proteins biologists are able to observe the location and distribution of cells and organelles.

The main purpose of image analysis in the context of biology is to characterize the cell state or its components. Early on, human eyes where the sole judge on this, but a more objective, scalable and reproducible method was required. This commonly entailed identifying regions of interest and calculating metrics that represent these regions, such as intensity distributions.

Morphological profiling is a technique where we generate a large number of shape nad intensity features for all objects, cells or otherwise. These features are fed into statistical or machine learning methods to identify patterns. One of its biggest applications is drug discovery, where scientists leverages microscopy's low acquisition cost and high throughput \citep{sealDecadeSystematicReview2024}. Morphological profiling serves not only as validation but also as information orthogonal to other high throughput methods, such as genomic sequencing.
\subsection{The current state of bioimage analysis}
\label{sec:org8f5b33d}
The most widely used software to process biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}. Its intended audience is experimental biologists with scant programmatic experience, for whom it has been a boon. In contrast, its usefulness greatly diminishes for computer-savvy scientists who fear no command line. For them, CellProfiler means friction in pipelines comprised of multiple tools. Its catering for manually-defined workflows unbefits low-customization high throughput analyses, which are prevalent in the age of generalist segmentation and foundational models.

Integrating a CellProfiler pipelines with other tools is a struggle. Building plug-ins to add functionality is time and effort-consuming challenges that require an understanding of its interfaces. Treating it as a black box and using its output precludes the addition of custom-made image filters. Lastly, CellProfiler depends on many number of packages from both Python and Java, amplifying the risk of dependency conflicts. Containers mitigate this problem, but not without increased complexity and caveats. To summarize, CellProfiler is an all-or-nothing tool, integrating other tools demands more work and time than most scientists can spare.

Object featurization is not exclusive of CellProfiler. For example, scikit-image can produce some overlapping features. Tools like ScaleFex \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} generate features, but they have limitations: First, their implementation is completely independent and thus cannot reproduce existing datasets, potentially impacting the biological interpretations. Moreover, a common portability problem is that these tools enforce a specific shape for the input data and require wrangling with the data before processing a single image. Some alternatives, such as ScaleFex, are designed for cloud usage first, and may neglect a significant number of cases where local compute suffices. We thus believe that there is the need for a tool aimed at generalist and high throughput problems not being solved by the current software ecosystem.
\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
The unprecedented evolution of deep learning methods in recent years has made available many different ways to process imaging data. Nevertheless, these are limited in their capacity to provide biological interpretation \citep{moenDeepLearningCellular2019}. Engineered features can also outperform deep learning ones under certain conditions \citep{tangMorphologicalProfilingDrug2024}. Even as deep learning features improve over time, they still require complex interpretability methods, even if they manage to be interpretable at all. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require data annotating. Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution data. Unlike deep learning-based features, whose interpretation requires complex methods like counterfactuals or attention maps, most engineered features have a mathematical definition that is faster to calculate and simpler to translate into biological concepts.

The intended audience for such a featurization tool would be, in addition to biologists, data scientists striving to achieve reproducible data analyses within a few lines of code. As a whole, engineered features still have the role of, at the minimum, complementing deep learning ones. The bioimage software ecosystems lacks a modular tool to featurize images into interpretable features with no compulsory graphical interfaces or cloud compute. Ideally it should also harmonize with other scientific libraries. With all this in mind we have developed cp\_measure.
\section{Extracting interpretable features}
\label{sec:org61842b5}
In cp\_measure, we isolated the CellProfiler measurements and categorize them based on their inputs: One object and one imaging channel (e.g., intensity, texture), one object and multiple channels (such as correlation features) and multiple objects (e.g., number of neighbors).

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
\caption{\label{fig:overview}Our tool produces features from images by using the local information in every region of interest. This can be used to featurize the pairwise combination of all the available channels. The resultant matrices represent the entirety of the experiment in a much more compressed way, and can be processed with standard machine and deep learning methods.}
\end{figure}

The cp\_measure features were based off the CellProfiler implementation, but we separated the actual functional components (i.e., math) from the user-facing and orchestration components. It improves reusability, testability and its makes it possible for more people to access and improve its internals in the future. We positioned it to become a stepping stone in extensibility and maintainability. By isolating and cleaning the implemented mathematics of CellProfiler we vastly reduce the amount of time and manual effort required to perform data analyses while retaining the features present in numerous datasets.

Our aim is to remain consistent with the current scientific python ecosystem. Its interface corresponds that of the widely used scikit-image \citep{waltScikitimageImageProcessing2014}. This greatly reduces the effort needed to integrate it in existing workflows and tools.

@@

We first performed the numerical validation of cp\_measure, relative to the original CellProfiler features. In preparation, we collected 300 images corresponding to 150 perturbations from the JUMP dataset. the most significant phenotypes for a given measurement each. To ensure that we are using identical object masks, we segmented these images to obtain the cells and nuclei using CellProfiler, providing object masks and their associated features. Next, we applied cp\_measure on these masks with the original images and mapped the features from cp\_measure to CelProfiler. Lastly, we calculated a linear fit for the matched features and calculated their \(R^2\) value.

@@

Figure \ref{fig:astrocytes} shows an example image and object masks alongside the Shapley values of a classifier trained on cp\_measure features. Our results showcased the axis length of the cell to be a major indicator of phenotypic effect, implying that cells became more elongated on their minor axis. While it was posisble to CellProfiler for this task, cp\_measure made it much faster and simpler to implement, specially since the segmentation masks were already provided, which is quite common in public datasets.
\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
A key advantage of providing these measurements as a standalone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment. The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality. However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment. In contrast, classical morphological measurements yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.

To demonstrate this utility, we integrated our cp\_measure-based feature extraction into the widely used spatial analysis library Squidpy (CITE). Being standalone allowed seamless incorporation into workflows powered by the robust SpatialData (CITE) framework underlying Squidpy. Because spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screenings -- up to approximately 100,000 cells-traditional software typically cannot process these large images without cropping, which introduces boundary artifacts. Leveraging the modular design of cp\_measure, we parallelized feature extraction at the single-cell level, streaming batches of cells across computational cores. This approach enables efficient computation even on large-scale datasets, a feat not achievable with standard CellProfiler software.

To further illustrate the value of morphological features, we evaluated their impact on cell-type prediction tasks using spatial transcriptomics data. This application is particularly compelling, as current spatial transcriptomics technologies typically produce matched histological images that remain largely underutilized beyond visualization. We analyzed two mouse brain datasets generated by Bruker Spatial's CosMx platform (CITE nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-mouse-brain-ffpe-dataset/). Each dataset comprises expression profiles for 960 genes and immunofluorescence images captured via five distinct fluorescent probes ('Histone', 'DNA', 'GFAP', 'G', 'rRNA'). Morphological features were extracted from these 5-channel images for both datasets. Subsequently, both gene expression and morphological data were preprocessed according to best practices established by Scanpy (CITE) and PyCytoMiner (CITE) respectively. We trained an XGBoost model to predict cell types on the larger dataset (48,556 cells; see Fig. XXX, panel XXX), comparing models using either gene expression alone or combined gene expression and morphological data. Model performance was assessed by predicting cell types in a smaller independent dataset (38,996 cells), using the F1-score metric stratified by cell type. Figure XXX (panel XXX) highlights the improved predictive accuracy obtained when morphological features are included. Importantly, this performance enhancement required no additional experimental effort, underscoring the benefit of employing cp\_measure beyond its traditional scope.

@@

Long-term, we envision cp\_measure can be the place to develop and distribute new measurements. While CellProfiler's measurements are already ubiquituous in bioimaging studies, the existing palette of measurements could be further extended to cover unexplored use-cases. We also see adding community-contributed measurements to better match the current questions scientists pose to imaging data.

>>>>

% Created 2025-05-19 Mon 21:10
% Intended LaTeX compiler: pdflatex
\documentclass{article}

@@

\begin{abstract}
Quantifying the contents of objects in images is a core challenge in biological imaging. The current tools require significant human intervention. Here we introduce our library cp\_measure, which provides programmatic access to the most widespread metrics to convert images and objects into features. We then demonstrate that the features are consistent to the standard ones and showcase tasks for which our tool is more suitable than the alternatives. cp\_measure opens the door to community-driven development and improvement of bioimage analysis metrics and pipelines, increasing the scaling capabilities, reproducibility and accessibility for computational and data scientists.
\end{abstract}
\section{Introduction}
\label{sec:org4c9ba67}
The field of biological imaging leverages data from cell microscopy images to obtain biological insights. Modern microscopes can capture a wide range of information, such as illumination and fluorescence. When coupled with fluorescence dyes or proteins biologists are able to observe the location and distribution of cells and organelles.

The main purpose of image analysis in biology is to characterize the cell state or its components. Early on, human eyes where the sole judge on this, but a more objective, scalable and reproducible method was required. This commonly entailed identifying regions of interest and calculating metrics that represent these regions, such as intensity distributions.

Morphological profiling is a technique where we generate an array of shape and intensity features for all objects, cells or otherwise. These features are fed into statistical or machine learning methods to identify patterns. One of its biggest applications is drug discovery, where scientists leverages microscopy's low acquisition cost and high throughput \citep{sealDecadeSystematicReview2024}. Morphological profiling serves not only as validation but also as information orthogonal to other high throughput methods, such as genomic sequencing.
\subsection{The current state of bioimage analysis}
\label{sec:org8f5b33d}
The most widely used software to process biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}. Its intended audience is experimental biologists with scant programmatic experience, for whom it has been a boon. In contrast, its usefulness greatly diminishes for computer-savvy scientists who fear no command line. For them, CellProfiler means friction in pipelines comprised of multiple tools. It caters for manually-defined workflows instead of low-customization high throughput analyses, which are becoming prevalent in the age of generalist segmentation and foundational models.

Integrating a CellProfiler pipelines with other tools is a struggle. Building plug-ins to add functionality is time and effort-consuming challenges that require an understanding of its interfaces. Treating it as a black box and using its output precludes the addition of custom-made image filters. Lastly, CellProfiler depends on many number of packages from both Python and Java, amplifying the risk of dependency conflicts. Containers mitigate this problem, but not without increased complexity and caveats. To summarize, CellProfiler is an all-or-nothing tool, whose extensibility and integration with other tools demands more time and effort than most scientists can spare.

Object featurization is not exclusive of CellProfiler. For example, scikit-image can produce some overlapping features \citep{waltScikitimageImageProcessing2014}. Tools like ScaleFex or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} can generate features, but they have limitations: First, their implementation is completely independent and thus cannot reproduce existing datasets, potentially impacting the biological interpretations. Moreover, a common problem is that these tools enforce a structure for the input data and require wrangling with the data before processing a single image. Some alternatives, such as ScaleFex, are designed for cloud usage first, and may neglect a significant number of cases where local compute suffices. We thus believe that there is the need for a tool aimed at generalist and high throughput problems not being solved by the current software ecosystem.
\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
The latest deep learning models can streamline image processing. Nevertheless, their capacity to provide biological interpretation is limited \citep{moenDeepLearningCellular2019}. In some cases, engineered features can even outperform learned ones \citep{tangMorphologicalProfilingDrug2024}. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require annotating data. Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution data. Unless a major change in neural networks occurs, they will still require more complex interpretability methods. Unlike learned features, whose interpretation requires complex methods like counterfactuals or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.

The intended audience for such a featurization tool would be, in addition to biologists, data scientists striving to achieve reproducible data analyses within a few lines of code. As a whole, engineered features still have the role of, at the minimum, complementing deep learning ones. The bioimage software ecosystems lacks a modular tool to convert images into interpretable features with no compulsory graphical interfaces or cloud compute. Ideally it should also harmonize with other scientific libraries. With all this in mind we have developed cp\_measure.
\section{Extracting interpretable features}
\label{sec:org61842b5}
We build cp\_measure by extracting the CellProfiler measurements from the code base and grouping them based on their inputs: One object and one imaging channel (e.g., intensity, texture), one object and multiple channels (e.g., Pearson correlation) and multiple objects (e.g., number of neighbours).

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
\caption{\label{fig:overview}cp\_measure generates features from images by using the local information in every region of interest. It is commonly used to featurise the pairwise combination of all the available channels and objects. The resultant matrices represent the entire experiment and can be studied using machine and deep learning methods.}
\end{figure}

By separating the numerical from the user-facing and orchestration components we improved reusability, testability and extensibility. We positioned it to become a stepping stone to make the morphological profiling numerical methods accessible to the data science community. Given that the numerical internals are more transparent, we reduced the amount of time and human effort required to perform data analyses, while still retaining the standard features already present in numerous datasets.

Our aim is to remain consistent with the current scientific Python ecosystem. Its interface corresponds that of the widely used scikit-image. This greatly reduces the effort needed to integrate it in existing workflows and tools.

@@

We first performed the numerical validation of cp\_measure, relative to the original CellProfiler features. For this we collected 300 images corresponding to 150 perturbations from the JUMP dataset, selecting the most significant phenotypes for a given measurement each. To ensure that we are using identical object masks, we segmented these images to obtain the cells and nuclei using CellProfiler, providing object masks and their associated features. Next, we applied cp\_measure on these masks with the original images and mapped the features from cp\_measure to CellProfiler. Lastly, we calculated a linear fit for the matched features and calculated their \(R^2\) value, indicating how well it fits a linear slope.

@@

Figure \ref{fig:astrocytes} shows an example image and object masks alongside the Shapley values of a classifier trained on cp\_measure features. Our results showcased the axis length of the cell to be a major indicator of phenotypic effect, implying that cells became more elongated on their minor axis. While it was possible to CellProfiler for this task, cp\_measure made it much faster and simpler to implement, specially since the segmentation masks were already provided, which is quite common in public datasets.
\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
A key advantage of providing these measurements as a standalone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment. The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality. However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment. In contrast, classical morphological measurements yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.

To demonstrate this utility, we integrated our cp\_measure-based feature extraction into the widely used spatial analysis library Squidpy (CITE). Being standalone allowed seamless incorporation into workflows powered by the robust SpatialData (CITE) framework underlying Squidpy. Because spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screenings -- up to approximately 100,000 cells-traditional software typically cannot process these large images without cropping, which introduces boundary artefacts. Leveraging the modular design of cp\_measure, we parallelized feature extraction at the single-cell level, streaming batches of cells across computational cores. This approach enables efficient computation even on large-scale datasets, a feat not achievable with standard CellProfiler software.

To further illustrate the value of morphological features, we evaluated their impact on cell-type prediction tasks using spatial transcriptomics data. This application is particularly compelling, as current spatial transcriptomics technologies typically produce matched histological images that remain largely underutilized beyond visualization. We analysed two mouse brain datasets generated by Bruker Spatial's CosMx platform (CITE nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-mouse-brain-ffpe-dataset/). Each dataset comprises expression profiles for 960 genes and immunofluorescence images captured via five distinct fluorescent probes ('Histone', 'DNA', 'GFAP', 'G', 'rRNA'). Morphological features were extracted from these 5-channel images for both datasets. Subsequently, both gene expression and morphological data were preprocessed according to best practices established by Scanpy (CITE) and Pycytominer (CITE) respectively. We trained an XGBoost model to predict cell types on the larger dataset (48,556 cells; see Fig. XXX, panel XXX), comparing models using either gene expression alone or combined gene expression and morphological data. Model performance was assessed by predicting cell types in a smaller independent dataset (38,996 cells), using the F1-score metric stratified by cell type. Figure XXX (panel XXX) highlights the improved predictive accuracy obtained when morphological features are included. Importantly, this performance enhancement required no additional experimental effort, underscoring the benefit of employing cp\_measure beyond its traditional scope.

@@

Long-term, we envision cp\_measure can be the place to develop and distribute new measurements. While CellProfiler's measurements are already ubiquitous in bioimaging studies, the existing palette of measurements could be further extended to cover unexplored use-cases. We also see adding community-contributed measurements to better match the current questions scientists pose to imaging data.
